---
title: "Assignment 3: Naive Bayes Classifier for Personal Loan Prediction"
author: "Nusrat Nasir Nimnee"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(e1071)
library(caret)
library(reshape2)
```

## Data Loading and Exploration

```{r load_data}
data <- read.csv("UniversalBank(1).csv")
str(data)
head(data)
summary(data)
```

## Data Preprocessing

```{r preprocess}
# Remove ID and ZIP Code
data <- data[, -c(1, 5)]

# Convert categorical variables to factors
data$Personal.Loan <- as.factor(data$Personal.Loan)
data$Online <- as.factor(data$Online)
data$CreditCard <- as.factor(data$CreditCard)

# Check for missing values
cat("Missing values:", sum(is.na(data)), "\n")
```

## Partition Data into Training (60%) and Validation (40%)

```{r partition}
set.seed(123)
trainIndex <- sample(1:nrow(data), 0.6 * nrow(data))
train <- data[trainIndex, ]
validation <- data[-trainIndex, ]

cat("Training set size:", nrow(train), "\n")
cat("Validation set size:", nrow(validation), "\n")
```

## A. Create Pivot Table with Online, CC, and Loan

```{r pivot_table_a}
# Create pivot table: Online (columns), CC (rows), Loan (secondary rows), Count (values)
pivot_table <- table(train$CreditCard, train$Online, train$Personal.Loan)
names(dimnames(pivot_table)) <- c("CreditCard", "Online", "Personal.Loan")

cat("Pivot Table - Count of customers:\n")
cat("Personal.Loan = 0 (Did not accept loan):\n")
print(pivot_table[,,"0"])
cat("\nPersonal.Loan = 1 (Accepted loan):\n")
print(pivot_table[,,"1"])

# Alternative flat table format
pivot_flat <- ftable(train$Personal.Loan, train$CreditCard, train$Online)
cat("\nFlat Pivot Table:\n")
print(pivot_flat)
```

## B. Probability P(Loan = 1 | CC = 1, Online = 1)

```{r prob_b}
# From pivot table: customers with CC=1 and Online=1
cc1_online1_loan1 <- pivot_table["1", "1", "1"]
cc1_online1_loan0 <- pivot_table["1", "1", "0"]
cc1_online1_total <- cc1_online1_loan1 + cc1_online1_loan0

prob_loan_given_cc_online <- cc1_online1_loan1 / cc1_online1_total

cat("Customers with CC=1 and Online=1 who accepted loan:", cc1_online1_loan1, "\n")
cat("Total customers with CC=1 and Online=1:", cc1_online1_total, "\n")
cat("P(Loan = 1 | CC = 1, Online = 1) =", round(prob_loan_given_cc_online, 4), "\n")
```

## C. Create Two Separate Pivot Tables

```{r pivot_tables_c}
# Pivot table: Loan (rows) by Online (columns)
pivot_loan_online <- table(train$Personal.Loan, train$Online)
rownames(pivot_loan_online) <- c("Loan=0", "Loan=1")
colnames(pivot_loan_online) <- c("Online=0", "Online=1")

cat("Pivot Table - Loan by Online:\n")
print(pivot_loan_online)

# Pivot table: Loan (rows) by CC (columns)
pivot_loan_cc <- table(train$Personal.Loan, train$CreditCard)
rownames(pivot_loan_cc) <- c("Loan=0", "Loan=1")
colnames(pivot_loan_cc) <- c("CC=0", "CC=1")

cat("\nPivot Table - Loan by CreditCard:\n")
print(pivot_loan_cc)
```

## D. Compute Required Probabilities

```{r probabilities_d}
# Extract counts from pivot tables
loan1_total <- sum(train$Personal.Loan == "1")
loan0_total <- sum(train$Personal.Loan == "0")
total_customers <- nrow(train)

# i. P(CC = 1 | Loan = 1)
cc1_loan1 <- pivot_loan_cc["Loan=1", "CC=1"]
p_cc1_given_loan1 <- cc1_loan1 / loan1_total

# ii. P(Online = 1 | Loan = 1)
online1_loan1 <- pivot_loan_online["Loan=1", "Online=1"]
p_online1_given_loan1 <- online1_loan1 / loan1_total

# iii. P(Loan = 1)
p_loan1 <- loan1_total / total_customers

# iv. P(CC = 1 | Loan = 0)
cc1_loan0 <- pivot_loan_cc["Loan=0", "CC=1"]
p_cc1_given_loan0 <- cc1_loan0 / loan0_total

# v. P(Online = 1 | Loan = 0)
online1_loan0 <- pivot_loan_online["Loan=0", "Online=1"]
p_online1_given_loan0 <- online1_loan0 / loan0_total

# vi. P(Loan = 0)
p_loan0 <- loan0_total / total_customers

cat("i.   P(CC = 1 | Loan = 1) =", round(p_cc1_given_loan1, 4), "\n")
cat("ii.  P(Online = 1 | Loan = 1) =", round(p_online1_given_loan1, 4), "\n")
cat("iii. P(Loan = 1) =", round(p_loan1, 4), "\n")
cat("iv.  P(CC = 1 | Loan = 0) =", round(p_cc1_given_loan0, 4), "\n")
cat("v.   P(Online = 1 | Loan = 0) =", round(p_online1_given_loan0, 4), "\n")
cat("vi.  P(Loan = 0) =", round(p_loan0, 4), "\n")
```

## E. Compute Naive Bayes Probability P(Loan = 1 | CC = 1, Online = 1)

```{r naive_bayes_e}
# Using Bayes theorem:
# P(Loan=1 | CC=1, Online=1) = P(CC=1 | Loan=1) * P(Online=1 | Loan=1) * P(Loan=1) / P(CC=1, Online=1)
# 
# For denominator: P(CC=1, Online=1) = P(CC=1, Online=1 | Loan=1)*P(Loan=1) + P(CC=1, Online=1 | Loan=0)*P(Loan=0)
# Assuming independence: P(CC=1, Online=1 | Loan=1) = P(CC=1|Loan=1) * P(Online=1|Loan=1)

# Numerator
numerator <- p_cc1_given_loan1 * p_online1_given_loan1 * p_loan1

# Denominator
prob_cc1_online1_given_loan1 <- p_cc1_given_loan1 * p_online1_given_loan1
prob_cc1_online1_given_loan0 <- p_cc1_given_loan0 * p_online1_given_loan0
denominator <- (prob_cc1_online1_given_loan1 * p_loan1) + (prob_cc1_online1_given_loan0 * p_loan0)

# Naive Bayes probability
nb_probability <- numerator / denominator

cat("Naive Bayes Calculation:\n")
cat("Numerator: P(CC=1|Loan=1) * P(Online=1|Loan=1) * P(Loan=1) =", round(numerator, 6), "\n")
cat("Denominator =", round(denominator, 6), "\n")
cat("\nP(Loan = 1 | CC = 1, Online = 1) =", round(nb_probability, 4), "\n")
```

## F. Compare with Pivot Table Result from (B)

```{r compare_f}
# Comparison
comparison_df <- data.frame(
  Method = c("Pivot Table (B)", "Naive Bayes (E)"),
  Probability = c(prob_loan_given_cc_online, nb_probability)
)
print(comparison_df)

cat("\nDifference:", round(abs(prob_loan_given_cc_online - nb_probability), 4), "\n")
cat("\nThe pivot table gives the exact probability from observed data, while Naive Bayes")
cat("\nassumes conditional independence. The pivot table is more accurate for this specific")
cat("\ncombination, but Naive Bayes generalizes better to new data.\n")
```

## G. Run Naive Bayes Model and Compare

```{r naive_bayes_model}
# Build Naive Bayes model using only Online and CreditCard predictors
nb_model <- naiveBayes(Personal.Loan ~ Online + CreditCard, data = train)
print(nb_model)

# Predict for customer with CC=1 and Online=1
test_customer <- data.frame(Online = factor("1", levels = c("0", "1")),
                            CreditCard = factor("1", levels = c("0", "1")))

prediction <- predict(nb_model, test_customer, type = "raw")
cat("\n\nModel prediction for customer with CC=1 and Online=1:\n")
cat("P(Loan = 0 | CC = 1, Online = 1) =", round(prediction[1], 4), "\n")
cat("P(Loan = 1 | CC = 1, Online = 1) =", round(prediction[2], 4), "\n")

cat("\nComparison:\n")
cat("Manually calculated (E):", round(nb_probability, 4), "\n")
cat("Model output:", round(prediction[2], 4), "\n")
cat("Difference:", round(abs(nb_probability - prediction[2]), 6), "\n\n")

cat("The values match closely, confirming our manual calculation.\n")
```

## Additional: Full Model with All Predictors

```{r full_model}
# Build full Naive Bayes model with all predictors
nb_full <- naiveBayes(Personal.Loan ~ ., data = train)

# Predictions on validation set
pred_val <- predict(nb_full, validation)

# Confusion matrix
conf_matrix <- confusionMatrix(pred_val, validation$Personal.Loan, positive = "1")
print(conf_matrix)

cat("\nModel Performance:\n")
cat("Accuracy:", round(conf_matrix$overall['Accuracy'], 4), "\n")
cat("Sensitivity (Recall):", round(conf_matrix$byClass['Sensitivity'], 4), "\n")
cat("Specificity:", round(conf_matrix$byClass['Specificity'], 4), "\n")
cat("Precision:", round(conf_matrix$byClass['Precision'], 4), "\n")
```



The Naive Bayes classifier successfully predicts personal loan acceptance with good accuracy. 
The manual calculations match the model output, confirming the correctness of the Naive Bayes 
approach. The key findings are:

1. Pivot Table Probability (B): More accurate for exact combinations seen in training data
2. Naive Bayes Probability (E): Uses independence assumption, generalizes better to unseen data
3. Model Performance: The full model achieves high accuracy on validation data
4. The conditional independence assumption works reasonably well for this dataset, though 
   there is a small difference between exact probability and Naive Bayes estimate.
