---
title: 'Assignment 1: Adult Income Analysis'
author: "Nusrat Nasir Nimnee"
date: "2025-09-05"
output:
  html_document:
    toc: true
    theme: united
  pdf_document:
    toc: true
  word_document:
    toc: true
---

# Introduction

This report analyzes the 'Adult Income' dataset to explore the factors that influence whether an individual earns over $50K a year.

```{r setup, include=FALSE}
# This chunk is for global setup options.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data Loading and Preparation
This dataset collected from kaggle dataset. 
Dataset Source: https://www.kaggle.com/datasets/serpilturanyksel/adult-income

```{r environment-setup}
# Clear the global environment
rm(list = ls())

# Load required packages
# install.packages("gtsummary") 
# install.packages("tidyverse")
library(tidyverse)
library(gtsummary)
```

```{r load-data}
# Load the dataset
# The dataset is expected to be in the same directory as this Rmd file.
Income_df <- read.csv("adult_income_data.csv")
```

# Exploratory Data Analysis

## Basic Summary

Here is a basic summary of the dataset.

```{r summary-base}
str(Income_df)
summary(Income_df)
```

## Detailed Summary Table

A more detailed and formatted summary using the `gtsummary` package, comparing income levels.

```{r summary-gtsummary}
Income_df %>%
  tbl_summary(by = salary) %>%
  add_p() %>%
  add_overall() %>%
  bold_labels()
```



# Data Transformation

Before visualization, let's transform at least one variable as required. We'll create a log transformation of hours per week to normalize the distribution.

```{r transform-data}
# Transform hours.per.week using log transformation
# Adding 1 to avoid log(0) issues
Income_df$log_hours_per_week <- log(Income_df$hours.per.week + 1)

# Show summary of original and transformed variable
cat("Original hours.per.week summary:\n")
summary(Income_df$hours.per.week)
cat("Log-transformed hours.per.week summary:\n")
summary(Income_df$log_hours_per_week)
```

# Data Visualization

Visualizing the data helps in understanding the relationships between different variables and the income level.

## Quantitative Variable Plot

```{r plot-quantitative}
# Plot of a quantitative variable: Age distribution
ggplot(Income_df, aes(x = age)) +
  geom_histogram(bins = 20, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Age", 
       x = "Age", 
       y = "Frequency") +
  theme_minimal()
```

## Scatterplot

```{r scatterplot}
# Scatterplot: Age vs Hours per week
ggplot(Income_df, aes(x = age, y = hours.per.week, color = salary)) +
  geom_point(alpha = 0.6) +
  labs(title = "Scatterplot: Age vs Hours per Week by Salary", 
       x = "Age", 
       y = "Hours per Week",
       color = "Salary") +
  theme_minimal()
```

## Additional Visualizations

```{r visualize-income}
# Distribution of Income
ggplot(Income_df, aes(x = salary)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Distribution of Income Levels", x = "Salary", y = "Count") +
  theme_minimal()
```

```{r visualize-transformed}
# Comparison of original vs transformed variable
par(mfrow = c(1, 2))
hist(Income_df$hours.per.week, main = "Original Hours per Week", xlab = "Hours per Week")
hist(Income_df$log_hours_per_week, main = "Log-transformed Hours per Week", xlab = "Log(Hours per Week + 1)")
```

# Conclusion

This report presents an exploratory data analysis of the adult income dataset. After loading data, several key insights were gathered through summary statistics and visualizations.

The analysis revealed the distribution of individuals earning above and below $50K. Visualizations of age, workclass, and other demographic factors against income level have highlighted potential predictors for higher income, such as age and type of employment. For instance, the proportion of individuals earning over $50K varies significantly across different workclasses.

These initial findings provide a solid foundation for understanding the dataset's characteristics and can inform future predictive modeling tasks.



