---
title: "Assignment 2: Predicting Personal Loan Acceptance using k-NN"
author: "Nusrat Nasir Nimnee"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
```

## Data Loading and Exploration

```{r load_data}
data <- read.csv("UniversalBank.csv")
str(data)
head(data)
summary(data)
```

## Data Preprocessing

```{r preprocess}
data$Personal.Loan <- as.factor(data$Personal.Loan)
data$Education <- as.factor(data$Education)
# Check for missing values
sum(is.na(data))
# Remove ID and ZIP Code as per instructions
data <- data %>% select(-ID, -ZIP.Code)
```

## Exploratory Data Analysis

```{r eda}
# Distribution of Personal Loan
ggplot(data, aes(x = Personal.Loan)) + 
  geom_bar(fill = "blue") + 
  labs(title = "Distribution of Personal Loan Acceptance")

# Summary of numeric variables
numeric_vars <- data %>% select_if(is.numeric)
summary(numeric_vars)
```

## Partition Data into Training (60%) and Validation (40%)

```{r partition}
set.seed(123)
trainIndex <- createDataPartition(data$Personal.Loan, p = 0.6, list = FALSE)
train <- data[trainIndex, ]
validation <- data[-trainIndex, ]
```

## k-NN Classification with k=1

```{r knn_k1}
set.seed(123)
# Train k-NN with k=1
model_k1 <- train(Personal.Loan ~ ., data = train, method = "knn", 
                  preProcess = c("center", "scale"), 
                  tuneGrid = data.frame(k = 1))
print(model_k1)
```

## Classify the Specific Customer with k=1

```{r classify_customer_k1}
# Customer data
customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Education = factor(2, levels = levels(data$Education)),  # Since Education_2=1
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1
)

# Predict
pred_customer_k1 <- predict(model_k1, customer)
print(pred_customer_k1)
```

## Choosing the Best k

```{r best_k}
set.seed(123)
# Train k-NN with different k values
model_best <- train(Personal.Loan ~ ., data = train, method = "knn", 
                    preProcess = c("center", "scale"), 
                    tuneLength = 10)
print(model_best)
best_k <- model_best$bestTune$k
print(paste("Best k:", best_k))
```

## Confusion Matrix for Validation with Best k

```{r conf_matrix_best}
pred_val <- predict(model_best, validation)
conf_matrix <- confusionMatrix(pred_val, validation$Personal.Loan)
print(conf_matrix)
```

## Classify the Customer with Best k

```{r classify_customer_best}
pred_customer_best <- predict(model_best, customer)
print(pred_customer_best)
```

## Repartition into Training (50%), Validation (30%), Test (20%)

```{r repartition}
set.seed(123)
trainIndex2 <- createDataPartition(data$Personal.Loan, p = 0.5, list = FALSE)
train2 <- data[trainIndex2, ]
remaining <- data[-trainIndex2, ]
valIndex <- createDataPartition(remaining$Personal.Loan, p = 0.6, list = FALSE)  # 30% of total
validation2 <- remaining[valIndex, ]
test <- remaining[-valIndex, ]
```

## Apply k-NN with Best k on New Partition

```{r knn_new_partition}
set.seed(123)
model_new <- train(Personal.Loan ~ ., data = train2, method = "knn", 
                   preProcess = c("center", "scale"), 
                   tuneGrid = data.frame(k = best_k))

# Predictions
pred_train <- predict(model_new, train2)
pred_val2 <- predict(model_new, validation2)
pred_test <- predict(model_new, test)

# Confusion Matrices
conf_train <- confusionMatrix(pred_train, train2$Personal.Loan)
conf_val <- confusionMatrix(pred_val2, validation2$Personal.Loan)
conf_test <- confusionMatrix(pred_test, test$Personal.Loan)

print("Training Confusion Matrix:")
print(conf_train)
print("Validation Confusion Matrix:")
print(conf_val)
print("Test Confusion Matrix:")
print(conf_test)
```

## Comments on Differences

The confusion matrices reveal that the training set typically shows higher accuracy due to the model fitting closely to the training data, which can indicate some overfitting. The validation and test sets, being unseen data, show slightly lower but more realistic performance metrics. The similarity between validation and test accuracies suggests good generalization of the model. Differences between sets are primarily due to random partitioning and the inherent variability in the data. The chosen k balances between underfitting (too high k, ignoring patterns) and overfitting (too low k, too sensitive to noise).