---
title: "Assignment 2: Predicting Personal Loan Acceptance using k-NN"
author: "Nusrat Nasir Nimnee"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
```

## Data Loading and Exploration

```{r load_data}
data <- read.csv("UniversalBank.csv")
str(data)
head(data)
summary(data)
```

## Data Preprocessing

```{r preprocess}
data$Personal.Loan <- as.factor(data$Personal.Loan)
data$Education <- as.factor(data$Education)
# Check for missing values
sum(is.na(data))
# Remove ID and ZIP Code as per instructions
data <- data %>% select(-ID, -ZIP.Code)

# Create dummy variables for Education manually
data$Education_1 <- ifelse(data$Education == 1, 1, 0)
data$Education_2 <- ifelse(data$Education == 2, 1, 0)
data$Education_3 <- ifelse(data$Education == 3, 1, 0)

# Remove the original Education variable
data <- data %>% select(-Education)

# Verify the dummy variables
cat("Dummy variables created successfully:\n")
cat("Education_1 :", sum(data$Education_1), "customers\n")
cat("Education_2 :", sum(data$Education_2), "customers\n") 
cat("Education_3 :", sum(data$Education_3), "customers\n")
```

## Partition Data into Training (60%) and Validation (40%)

```{r partition}
set.seed(123)
trainIndex <- createDataPartition(data$Personal.Loan, p = 0.6, list = FALSE)
train <- data[trainIndex, ]
validation <- data[-trainIndex, ]
```

## k-NN Classification with k=1

```{r knn_k1}
set.seed(123)
# Train k-NN with k=1
model_k1 <- train(Personal.Loan ~ ., data = train, method = "knn", 
                  preProcess = c("center", "scale"), 
                  tuneGrid = data.frame(k = 1))
print(model_k1)
```

## Classify the Specific Customer with k=1

```{r classify_customer_k1}
# Customer data with dummy variables as specified in assignment:
# Age = 40, Experience = 10, Income = 84, Family = 2, CCAvg = 2, 
# Education_1 = 0, Education_2 = 1, Education_3 = 0, Mortgage = 0, 
# Securities Account = 0, CD Account = 0, Online = 1, Credit Card = 1
customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Education_1 = 0,  
  Education_2 = 1, 
  Education_3 = 0, 
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1
)

# Predict
pred_customer_k1 <- predict(model_k1, customer)
cat("Customer classification with k=1:", as.character(pred_customer_k1))
if(pred_customer_k1 == 1) {
  cat("\nPrediction: Customer will ACCEPT the personal loan offer")
} else {
  cat("\nPrediction: Customer will REJECT the personal loan offer")
}
```

## Choosing the Best k

```{r best_k}
set.seed(123)
# Train k-NN with k values from 1 to 20 for comprehensive evaluation
# Using cross-validation to find truly optimal k (not just training accuracy)
model_best <- train(Personal.Loan ~ ., data = train, method = "knn", 
                    preProcess = c("center", "scale"), 
                    tuneGrid = data.frame(k = 1:20),
                    trControl = trainControl(method = "cv", number = 5))
print(model_best)
best_k <- model_best$bestTune$k
cat("Best k:", best_k, "\n")
cat("Best cross-validated accuracy:", round(max(model_best$results$Accuracy), 4), "\n")

# Create a comprehensive plot to visualize k selection
library(ggplot2)
results_df <- model_best$results

ggplot(results_df, aes(x = k)) + 
  geom_line(aes(y = Accuracy, color = "CV Accuracy"), size = 1.2) +
  geom_point(aes(y = Accuracy, color = "CV Accuracy"), size = 3) +
  geom_line(aes(y = Kappa, color = "Kappa"), size = 1.2) +
  geom_point(aes(y = Kappa, color = "Kappa"), size = 3) +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = best_k + 1, y = max(results_df$Accuracy) - 0.02, 
           label = paste("Best k =", best_k), color = "red", size = 4) +
  labs(title = "k-NN Model Performance: Finding Optimal k (5-Fold CV)",
       subtitle = "Cross-validation prevents overfitting and finds truly optimal k",
       x = "Number of Neighbors (k)", 
       y = "Performance Metric",
       color = "Metric") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12)) +
  scale_x_continuous(breaks = 1:20)

# Show detailed comparison
cat("\nDetailed k comparison:\n")
print(results_df[, c("k", "Accuracy", "Kappa")])
```

## Confusion Matrix for Validation with Best k

```{r conf_matrix_best}
pred_val <- predict(model_best, validation)
conf_matrix <- confusionMatrix(pred_val, validation$Personal.Loan)
print(conf_matrix)
```

## Classify the Customer with Best k

```{r classify_customer_best}
# Same customer data with optimal k
pred_customer_best <- predict(model_best, customer)
cat("Customer classification with best k (k =", best_k, "):", as.character(pred_customer_best))
if(pred_customer_best == 1) {
  cat("\nPrediction: Customer will ACCEPT the personal loan offer")
} else {
  cat("\nPrediction: Customer will REJECT the personal loan offer")
}
```

## Repartition into Training (50%), Validation (30%), Test (20%)

```{r repartition}
set.seed(123)
trainIndex2 <- createDataPartition(data$Personal.Loan, p = 0.5, list = FALSE)
train2 <- data[trainIndex2, ]
remaining <- data[-trainIndex2, ]
valIndex <- createDataPartition(remaining$Personal.Loan, p = 0.6, list = FALSE)  # 30% of total
validation2 <- remaining[valIndex, ]
test <- remaining[-valIndex, ]
```

## Apply k-NN with Best k on New Partition

```{r knn_new_partition}
set.seed(123)
model_new <- train(Personal.Loan ~ ., data = train2, method = "knn", 
                   preProcess = c("center", "scale"), 
                   tuneGrid = data.frame(k = best_k))

# Predictions
pred_train <- predict(model_new, train2)
pred_val2 <- predict(model_new, validation2)
pred_test <- predict(model_new, test)

# Confusion Matrices
conf_train <- confusionMatrix(pred_train, train2$Personal.Loan)
conf_val <- confusionMatrix(pred_val2, validation2$Personal.Loan)
conf_test <- confusionMatrix(pred_test, test$Personal.Loan)

print("Training Confusion Matrix:")
print(conf_train)
print("Validation Confusion Matrix:")
print(conf_val)
print("Test Confusion Matrix:")
print(conf_test)
```

## Comments on Differences

The confusion matrices show different performance across the three datasets. The training set has the highest accuracy (97.48%) because the model was trained on this data. The validation and test sets show similar accuracy (96.60% each), which is good for model generalization.

The training set shows some overfitting with higher Kappa value (0.8369) compared to validation (0.7787) and test (0.7698). However, the difference is not very large, suggesting the chosen k-value works well.

All datasets show high sensitivity (>99%) for identifying customers who will not accept loans. However, specificity is lower (65-75%), meaning the model has difficulty identifying customers who will accept loans. This happens because most customers (90.4%) do not accept loans, creating class imbalance.

The similar performance between validation and test sets indicates good model stability. The model generalizes well to new data without significant overfitting. The chosen k balances between overfitting (low k) and underfitting (high k), providing reasonable performance for loan prediction.