---
title: "Assignment 4: Cluster Analysis of Pharmaceutical Companies"
author: "Nusrat Nasir Nimnee"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
# Set global options for code chunks
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### **Purpose**
The purpose of this assignment is to use the k-Means clustering algorithm to explore and understand the structure of the pharmaceutical industry by analyzing financial data from 21 firms.

### **1. Data Loading and Preparation**

First, we will load the necessary libraries and import the data from the `Pharmaceuticals.csv` file.

```{r load_data}
# Load necessary libraries
library(tidyverse)    # For data manipulation and visualization
library(cluster)      # For clustering algorithms
library(factoextra)   # For visualizing clustering results

# Load the dataset
pharma_data <- read.csv("Pharmaceuticals.csv")
```

### **a) Cluster Analysis and Justification of Choices**

In this step, we will conduct the cluster analysis using only the numerical variables (1 to 9) and justify the choices made.

#### **Data Pre-processing: Variable Selection and Scaling**

As per the assignment directions, we will use the first nine numerical financial variables: `Market_Cap`, `Beta`, `PE_Ratio`, `ROE`, `ROA`, `Asset_Turnover`, `Leverage`, `Rev_Growth`, and `Net_Profit_Margin`.

**Justification for Scaling:** The k-Means algorithm is based on Euclidean distance, which is sensitive to the scale of variables. For instance, `Market_Cap` (in billions) has a much larger scale than `Beta`. If we do not scale the data, variables with larger scales will disproportionately influence the clustering result. Therefore, we will **standardize** the data (to have a mean of 0 and a standard deviation of 1) to ensure all variables contribute equally to the analysis.

```{r preprocessing}
# Select only the numerical variables for clustering (columns 3 to 11)
pharma_numeric <- pharma_data[, 3:11]

# Set company names as row names for easier identification
rownames(pharma_numeric) <- pharma_data$Name

# Scale (standardize) the data
pharma_scaled <- scale(pharma_numeric)

# View the first few rows of the scaled data
head(pharma_scaled)
```

#### **Determining the Optimal Number of Clusters (k)**

To choose an appropriate number of clusters, `k`, we will use two data-driven methods: the **Elbow Method** and the **Average Silhouette Method**.

**Justification for Choice of k:** These methods provide a systematic way to determine the optimal `k` by evaluating the quality of the clustering for different values of `k`. This avoids arbitrary selection and leads to more meaningful clusters.

```{r find_optimal_k}
# Determine optimal k using the Elbow Method
fviz_nbclust(pharma_scaled, kmeans, method = "wss") +
  labs(subtitle = "Elbow Method")

# Determine optimal k using the Silhouette Method
fviz_nbclust(pharma_scaled, kmeans, method = "silhouette") +
  labs(subtitle = "Silhouette Method")
```

**Observation:** The **Elbow Method** shows a distinct "elbow" at **k=3**, after which the decrease in the total within-cluster sum of squares (WSS) becomes less pronounced. 

The **Silhouette Method** indicates **k=5** yields the highest average silhouette width, **but k=3** also shows a strong silhouette value and is supported by the Elbow Method and suitability for a small dataset.

#### **Performing k-Means Clustering**

Now, we will run the k-Means algorithm with `k=3`. We use `set.seed` for reproducibility and `nstart = 25` to ensure the algorithm finds a stable and optimal solution.

```{r perform_kmeans}
# Set a seed for reproducibility
set.seed(123)

# Perform k-Means clustering with k=3
final_clusters <- kmeans(pharma_scaled, centers = 3, nstart = 25)

# Print the clustering results
print(final_clusters)

# Visualize the clusters
fviz_cluster(final_clusters, data = pharma_scaled,
             palette = c("#2E9FDF", "#00AFBB", "#E7B800"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
)
```

### **b) Interpretation of the Clusters**

To interpret the clusters, we will analyze the mean values of the original (unscaled) numerical variables for each cluster. This allows us to understand the real-world characteristics of each group.

```{r interpret_clusters}
# Add the cluster assignment to the original numeric data
pharma_numeric_clustered <- pharma_numeric %>%
  mutate(Cluster = final_clusters$cluster)

# Calculate the mean of each variable for each cluster
cluster_summary <- pharma_numeric_clustered %>%
  group_by(Cluster) %>%
  summarise_all(mean)

# Print the summary table
print.data.frame(cluster_summary)
```

**Cluster Interpretations (Corrected):**

Based on the summary table above, we can interpret the clusters as follows:

*   **Cluster 1: Stable, Highly-Valued Competitors.** These firms exhibit moderate `Market_Cap` ($21.8B) and `ROE` (11.3). Their most defining feature is the **highest average PE_Ratio (46.9)**, indicating strong investor confidence in their future, despite having the lowest revenue growth and net profit margin in this dataset.

*   **Cluster 2: Large, Profitable Blue-Chip Firms.** This group represents the industry giants, with the **highest average Market_Cap ($97.1B)**, the **highest ROE (35.7)**, and the **highest Net_Profit_Margin (20.2%)**. Their financial performance is exceptionally strong and stable.

*   **Cluster 3: High-Growth, High-Risk Emerging Firms.** This cluster contains the smallest firms, with the **lowest average Market_Cap ($9.2B)**. However, they compensate with the **highest Revenue Growth (23.5%)**. Their high-risk nature is evident from the **highest average Leverage (1.25)**.

### **c) Pattern with Respect to Non-clustering Variables**

Now, we examine if there is a pattern in the clusters with respect to the categorical variables (10-12) that were not used in forming the clusters.

```{r pattern_analysis}
# Add cluster assignments to the categorical variables
pharma_full_clustered <- pharma_data %>%
  select(Name, Median_Recommendation, Location, Exchange) %>%
  mutate(Cluster = final_clusters$cluster)

# Pattern by Median Recommendation
cat("Cluster distribution by Median Recommendation:\n")
table(pharma_full_clustered$Cluster, pharma_full_clustered$Median_Recommendation)

# Pattern by Location
cat("\nCluster distribution by Location:\n")
table(pharma_full_clustered$Cluster, pharma_full_clustered$Location)

# Pattern by Stock Exchange
cat("\nCluster distribution by Stock Exchange:\n")
table(pharma_full_clustered$Cluster, pharma_full_clustered$Exchange)
```

**Observations:**

*   **Median Recommendation:** Cluster 2 (Large Blue-Chips) firms are mostly rated as `Hold` or `Moderate Buy`. The other two clusters have more mixed recommendations.
*   **Location:** Cluster 2 is dominated by firms headquartered in the **US** and **UK**. Most firms in Cluster 3 (High-Growth) are based in the **US**. Cluster 1 (Stable Competitors) is more geographically diverse.
*   **Exchange:** Almost all firms in Cluster 1 and 2 are listed on the **NYSE**. The firms in Cluster 3 are listed on various exchanges, which is consistent with smaller, emerging companies.

### **d) Provide an Appropriate Name for Each Cluster**

Based on the combined financial and categorical analysis, we can provide the following descriptive names for each cluster:

*   **Cluster 1: Stable International Competitors**
    *   **Reason:** This group consists of well-established, geographically diverse firms with stable performance. Their high PE ratios indicate strong investor confidence in their long-term value, even with lower current profitability.

*   **Cluster 2: Global Pharma Giants**
    *   **Reason:** This group includes companies with massive market caps, superior profitability, and a dominant presence in major markets (US/UK) and on the premier stock exchange (NYSE).

*   **Cluster 3: Dynamic Growth Seekers**
    *   **Reason:** These are smaller, riskier firms characterized by their aggressive revenue growth. They are dynamic and likely focused on innovation or niche markets to challenge the established giants.
```